### 2013-06-02 paper_draft.tex
Added FoldIt to table.
small changes to crowdsourcing + added citations

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Genome-wide association studies are widely used to correlate phenotypic traits with genetic variants. These studies usually compare the genetic variation between two groups to single out certain Single Nucleotide Polymorphisms (SNPs) that are linked to a phenotypic variation in one of the groups. However, it is necessary to have a large enough sample size to find statistically significant correlations. Direct-To-Consumer (DTC) genetic testing can supply additional data: DTC-companies offer the analysis of a large amount of SNPs for an individual at low cost without the need to consult a physician or geneticist. Over 100,000 people have already been genotyped through Direct-To-Consumer genetic testing companies. However, this data is not public for a variety of reasons and thus cannot be used in research. It seems reasonable to create a central open data repository for such data, but it was previously unknown if and how people would submit their data to such a repository. Here 
%we present a survey which evaluates whether people are willing to publicly share their genetic information. In the light of those results 
we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data at no cost along with phenotypic information. Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a 
%valuable 
resource for a wide area of studies, including Genome-Wide Association S
tudies. openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr

@@

\subsection*{Other crowdsourcing platforms}
Projects similar to openSNP are the SNPedia, the Personal Genome Project and PatientsLikeMe.com (see table \ref{tab:platforms} for an overview). The focus of the SNPedia is the aggregation and summary of primary scientific literature on SNPs. The project uses a Wiki to store and display the data collected by volunteers contributing to the project. The data is mainly organized by the unique Rs-ID, as given by dbSNP. If Rs-IDs are missing the identifiers given by the DTC testing companies may be used, similar to the way openSNP stores the data. For individual SNPs pages may list scientific literature and summaries on the found impact can be given. As those pages are largely created manually and not automatically through database access these summaries may not be complete. openSNP utilizes the SNPedia in that it crawls their data for SNPs, the summary of the impact and the magnitude a SNP has. While they offer a page listing download-URLs the SNPedia does not offer any uploading capabilities for genetic data and has no APIs to easily access the different data sets.

The Personal Genome Project (PGP) has its focus on collecting and hosting genetic as well as phenotypic data. Unlike openSNP they do not offer a completely open enrollment. For each participant of the PGP eligibility has to be established and participants have to give IRB approved informed consent. This allows for an easier re-use of the data, but at the same time makes it impossible for many people to enroll (e.g. non-US citizens). Depending on the specific use one has for the data this might be preferable to the open approach openSNP takes. While the PGP stores genotyping data as well as exome and genome data sets it is currently impossible to access this data through an API, instead data has to be manually extracted from their database. The annotation database of the PGP is not aimed at delivering specific publications but instead focuses more on specific traits (IS THAT RIGHT?!). The  annotation data stored by the PGP is incorporated into openSNP as well.

PatientsLikeMe is a community for patients with life-changing illnesses to track and share the development of their illness with other patients with similar illnesses \cite{Wicks2010}.  This helps patients in understanding their diseases better - 72\% of surveyed participants found the site "moderately" or "very helpful", for example when it comes to starting a new medication (37\% found the site helpful), or when it comes to changing the medication (27\%). Some subsets of data stored in PatientsLikeMe are open to the public and has been shown to be useful for research, for example in Multiple Sclerosis \cite{Bove2013}.

There are some projects that use gamification to let players crowdsource scientific data. For example, FoldIt is a puzzle game that lets players fold protein structures in order to achieve optimal structures. Players of Foldit have been able to identify protein structure and were even able to improve the activity of existing proteins structures \cite{Eiben2012}.
%COMPARE MORE. DO A TABLE. SHOULD THIS MAYBE BE PUT IN THE DISCUSSION INSTEAD OF RESULTS?

% I think discussion is better - we might even change the comparison a bit and put it into introduction, by just comparing SNPedia, PGP etc.

@@

%As we have found in our survey on sharing such results (see supplementary methods), many 
DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information. There is a variety of ethical and privacy implications when it comes to DTC genetic testing \cite{Caulfield2011,Joh2011}.

%Our survey has shown that 
People are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments or advertisers might misuse the information. Policy makers start to react to those changes by introducing laws like the 
\textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of
widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data.

@@

With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic 
data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to freely use in their studies.
Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. 
While there is not enough data uploaded to perform a statistically sound GWAS yet, this will be possible in the future, as user numbers continue to rise. By including the option of uploading exome data sets the platform already is capable of adjusting for changes in the type of data generated by DTC genetic testing.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}

@@

A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in Figure \ref{Figure4_label}. The source code of openSNP is 
published under the MIT license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
We thank Dr. Manuel Corpas and Prof. David Edwards for constructive advice in grammar, spelling and structure of this study. Further thanks go to Samantha Clark and Dan Bolser for providing valuable feedback and feature ideas during the development and Thomas Down, author of \emph{BioDalliance}, and Rafael Jimenez, author of \emph{MyKaryoView}, for their support on implementing the Distributed Annotation System and the genome browser. For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams. We would especially like to thank the users of openSNP.org for their participation, their constructive criticism and bug-finding abilities and especially for sharing their genotyping and phenotype data.

@@

\section*{Tables}
\begin{table}
\caption{Comparison of crowdsourced genetics platforms.}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
Name & Provides annotation & Provides data & Open enrollment & API & License \\ 
\hline
SNPedia & \checkmark & \xmark & N/A & \checkmark & CC-NC-SA 3.0 \\ 
\hline
PGP & \checkmark & \checkmark & \xmark & \xmark & CC-BY 1.0, CC0 \\
\hline
PatientsLikeMe & \xmark & \checkmark & \checkmark & \xmark & Closed \& CC-BY-SA 3.0 \\
\hline
openSNP & \checkmark & \checkmark & \checkmark & \checkmark & CC-BY 3.0 \\
\hline
\end{tabular}
\label{tab:platforms}
\end{table}
%\begin{table}
%\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped. }  
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean \emph{Genotyped} & Mean \emph{Not Genotyped} & Mean Difference & Standard Error\\ 
%\hline
%\textbf{Motivation for sharing data in participants who are already genotyped} & & & & \\
%\hline 
%... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ \hline % checked
%... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\ \hline % checked
%... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline % checked
%\textbf{Motivation for not sharing in participants who are not planning to get genotyped} & & & & \\ \hline
%... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\ \hline % checked
%... breach of privacy & 3.01 & 3.68  & 0.666 & 0.211 \\ \hline % checked
%... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ \hline % checked
%... negative consequences for family members & 2.93 & 3.57  & 0.639 & 0.197 \\ \hline 
%\end{tabular}
%\label{tab:motivations1}
%\end{table}
%
%\begin{table}
%\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean {Sharing} & Mean {Not Sharing} & Mean Difference & Standard Error \\ 
%\hline
%\textbf{Motivation for sharing genotypings in participants who would share} & & & & \\
%\hline
%... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\ \hline % checked
%... want to help science & 4.70 & 3.13 & 1.57 & 0.199 \\ \hline % checked
%... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline % checked
%\textbf{Motivation for sharing genotypings in participants who would not share} & & & & \\
%\hline
%... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\ \hline % checked
%... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\ \hline % checked
%... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline % checked
%\end{tabular}
%\label{tab:motivations2}
%\end{table}
%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}
% \end{table}

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Genome-Wide Association Studies are widely used to correlate phenotypic traits with genetic variants. These studies usually compare the genetic variation between two groups to single out certain Single Nucleotide Polymorphisms (SNPs) that are linked to a phenotypic variation in one of the groups. However, it is necessary to have a large enough sample size to find statistically significant correlations. Direct-To-Consumer (DTC) genetic testing can supply additional data: DTC-companies offer the analysis of a large amount of SNPs for an individual at low cost without the need to consult a physician or geneticist. Over 100,000 people have already been genotyped through Direct-To-Consumer genetic testing companies. However, this data is not public for a variety of reasons and thus cannot be used in research. It seems reasonable to create a central open data repository for such data, but it was previously unknown if and how people would submit their data to such a repository. Here 
%we present a survey which evaluates whether people are willing to publicly share their genetic information. In the light of those results 
we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish their genetic data at no cost along with phenotypic information. Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a 
%valuable 
resource for a wide area of studies, including Genome-Wide Association S
tudies. openSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr

@@

\subsection*{Other crowdsourcing platforms}
Projects similar to openSNP are the SNPedia, the Personal Genome Project and PatientsLikeMe.com (see table \ref{tab:platforms} for an overview). The focus of the SNPedia is the aggregation and summary of primary scientific literature on SNPs. The project uses a Wiki to store and display the data collected by volunteers contributing to the project. The data is mainly organized by the unique Rs-ID, as given by dbSNP. If Rs-IDs are missing the identifiers given by the DTC testing companies may be used, similar to the way openSNP stores the data. For individual SNPs pages may list scientific literature and summaries on the found impact can be given. As those pages are largely created manually and not automatically through database access these summaries may not be complete. openSNP utilizes the SNPedia in that it crawls their data for SNPs, the summary of the impact and the magnitude a SNP has. While they offer a page listing download-URLs the SNPedia does not offer any uploading capabilities for genetic data and has no APIs to easily access SNPs or data subsets in the different data sets. Similarly there is no way for users of SNPedia to share their phenotypes in a machine readable format.

The Personal Genome Project (PGP) has its focus on collecting and hosting genetic as well as phenotypic data. Unlike openSNP they do not offer a completely open enrollment. For each participant of the PGP eligibility has to be established and participants have to give IRB approved informed consent. This allows for an easier re-use of the data, but at the same time makes it impossible for many people to enroll (e.g. non-US citizens). Depending on the specific use one has for the data this might be preferable to the open approach openSNP takes. While the PGP stores genotyping data as well as exome and genome data sets it is currently impossible to access this data through an API, instead data has to be manually extracted from their database. The annotation database of the PGP is not aimed at delivering specific publications but instead focuses more on specific traits. The annotation data stored by the PGP is incorporated into openSNP as well.

PatientsLikeMe is a community for patients with life-changing illnesses to track and share the development of their illness with other patients with similar illnesses \cite{Wicks2010}.  This helps patients in understanding their diseases better - 72\% of surveyed participants found the site "moderately" or "very helpful", for example when it comes to starting a new medication (37\% found the site helpful), or when it comes to changing the medication (27\%). Some subsets of data stored in PatientsLikeMe are open to the public and has been shown to be useful for research, for example in Multiple Sclerosis \cite{Bove2013}. Alternatively access to the data they store can be licensed by researchers for a fee.

There are some projects that use gamification to let players work with crowdsourced scientific data. For example, FoldIt is a puzzle game that lets players fold protein structures in order to achieve optimal structures. Players of Foldit have been able to identify protein structure and were even able to improve the activity of existing proteins structures \cite{Eiben2012}. Another example is Galaxy Zoo \cite{GalaxyZoo}, which allows everyone to perform classification tasks for galaxies based on images collected by the Sloan Digital Sky Survey.

Unlike PatientsLikeMe, the PGP, or openSNP - which give the task of collecting the data into the hands of the crowd - Foldit and Galaxy Zoo limit themselves to analyzing data which was previously collected by scientists.

%add another game here?

%COMPARE MORE. DO A TABLE. SHOULD THIS MAYBE BE PUT IN THE DISCUSSION INSTEAD OF RESULTS?

@@

%As we have found in our survey on sharing such results (see supplementary methods), many 
DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information (CITATION NEEDED). There is a variety of ethical and privacy implications when it comes to DTC genetic testing \cite{Caulfield2011,Joh2011}.

%Our survey has shown that 
People are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments or advertisers might misuse the information (CITATION NEEDED). Policy makers start to react to those changes by introducing laws like the 
\textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of
widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data (CITATION NEEDED).

@@

With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic 
data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to freely use in their studies.
Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. 
While there is not enough data uploaded to perform a statistically sound GWAS yet, this will be possible in the future, as user numbers continue to rise. By including the option of uploading exome data sets the platform already is capable of adjusting for changes in the type of data generated by DTC genetic testing. Future improvements made on openSNP will address interoperability with other platforms and tools in Personal Genomics, amongst others: The standardization of phenotypes, the inclusion of further annotation sources and support for a wider range of data sets, including full genome data.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}
\subsection*{Ethics Statement}
The director of the ethics commission of the Goethe University Frankfurt am Main, Germany teaching hospital confirmed that this study does not fall within their remit.

@@

A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in Figure \ref{Figure4_label}. The source code of openSNP is 
published under the MIT license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
We thank Dr. Manuel Corpas and Prof. David Edwards for constructive advice in grammar, spelling and structure of this study. Further thanks go to Samantha Clark, Fabian Zimmer and Dan Bolser for providing valuable feedback and feature ideas during the development and Thomas Down, author of \emph{BioDalliance}, and Rafael Jimenez, author of \emph{MyKaryoView}, for their support on implementing the Distributed Annotation System and the genome browser. For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams. We would especially like to thank the users of openSNP.org for their participation, their constructive criticism and bug-finding abilities and especially for sharing their genotyping and phenotype data.

@@

\section*{Tables}
\begin{table}
\caption{Comparison of crowdsourced genetics platforms.}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
Name & Provides annotation & Provides data & Open enrollment & API & License \\ 
\hline
SNPedia & \checkmark & \xmark & N/A & \checkmark & CC-NC-SA 3.0 \\ 
\hline
PGP & \checkmark & \checkmark & \xmark & \xmark & CC-BY 1.0, CC0 \\
\hline
PatientsLikeMe & \xmark & \checkmark & \checkmark & \xmark & Closed, CC-BY-SA 3.0 \\
\hline
openSNP & \checkmark & \checkmark & \checkmark & \checkmark & CC-BY 3.0 \\
\hline
\end{tabular}
\label{tab:platforms}
\end{table}
%\begin{table}
%\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped. }  
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean \emph{Genotyped} & Mean \emph{Not Genotyped} & Mean Difference & Standard Error\\ 
%\hline
%\textbf{Motivation for sharing data in participants who are already genotyped} & & & & \\
%\hline 
%... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ \hline % checked
%... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\ \hline % checked
%... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline % checked
%\textbf{Motivation for not sharing in participants who are not planning to get genotyped} & & & & \\ \hline
%... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\ \hline % checked
%... breach of privacy & 3.01 & 3.68  & 0.666 & 0.211 \\ \hline % checked
%... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ \hline % checked
%... negative consequences for family members & 2.93 & 3.57  & 0.639 & 0.197 \\ \hline 
%\end{tabular}
%\label{tab:motivations1}
%\end{table}
%
%\begin{table}
%\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean {Sharing} & Mean {Not Sharing} & Mean Difference & Standard Error \\ 
%\hline
%\textbf{Motivation for sharing genotypings in participants who would share} & & & & \\
%\hline
%... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\ \hline % checked
%... want to help science & 4.70 & 3.13 & 1.57 & 0.199 \\ \hline % checked
%... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline % checked
%\textbf{Motivation for sharing genotypings in participants who would not share} & & & & \\
%\hline
%... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\ \hline % checked
%... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\ \hline % checked
%... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline % checked
%\end{tabular}
%\label{tab:motivations2}
%\end{table}
%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}
% \end{table}
